---
title: "heirarchical_clustering"
author: "Maya Spaur"
date: "3/21/2020"
output: html_document
---
#code to read in and merge datasets copied from school_code Rmd


```{r setup}
library(tidyverse)
library(magrittr)
library(pastecs)
library(psych)
library(readxl)
library(lubridate)
library(dendextend)
library(ggdendro)
library(factoextra)
library(reshape2)
library(pals)
library(patchwork)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))

# Misbath: A note on merging
# My understanding is that the DBN associated with each school is made up of 3 elements
# - the geographical district number (2 digits)
# - the borough code (a capital letter, ie K is for Brooklyn)
# - the school/building number (3 digits)

```

### Demographics
```{r, message=FALSE}
demo <- read_csv("./school_data/demographics.csv") %>% 
  janitor::clean_names() %>% 
  filter(year == "2016-17") %>% 
  mutate(school_year = year) %>% 
  select(-year) %>%
  mutate(school_name = toupper(school_name)) %>%
  drop_na(school_name)

#Removing the % sign from "percent_" columns for future analyses 
cols = c(19, 21, 23, 25, 27, 29, 31, 33, 35, 37, 38)
demo[,cols] %<>% lapply(function(x) parse_number(x))

#The Economic Need Index (ENI) estimates the percentage of students facing economic hardship
#I also removed the % sign from that column, but need to keep in mind that it's a percentage 

#names(demo)
#str(demo)
#view(demo)

```

### Lead data
```{r, message=FALSE}
lead <- read_csv("./school_data/2017_water_lead.csv") %>% 
  janitor::clean_names() %>% 
  mutate(number_of_elevated_samples = as.numeric(number_of_elevated_samples),
         number_of_samples_tested = as.numeric(number_of_samples_tested),
         percent_elevated = 
           round((number_of_elevated_samples / number_of_samples_tested)*100, 2),
         geographical_district = as.character(geographical_district)) %>% 
#Adding leading 0 for district numbers while always having 2 digits 
  mutate(geographical_district = 
           str_pad(geographical_district, width=2, side="left", pad="0")) %>% 
  mutate(dbn = str_c(geographical_district, building_code)) %>% 
  drop_na(elevated_result) 
  
#summary(lead)
#names(lead)
#str(lead)
#view(lead)

```


### Scores
```{r, message=FALSE}
ela_scores <- read_csv("./school_data/ela_scores.csv") %>% 
  janitor::clean_names() %>% 
  filter(year %in% c("2016", "2017")) %>% 
  arrange(dbn, year) %>% 
  rename(mean_ela_score = "mean_scale_score", number_tested_ela = "number_tested") %>% 
  select(dbn:mean_ela_score)

#names(ela_scores)
#str(ela_scores)
#view(ela_scores)

math_scores <- read_csv("./school_data/math_scores.csv") %>% 
  janitor::clean_names() %>% 
  filter(year %in% c("2016", "2017")) %>%
  arrange(dbn, year) %>%
  rename(mean_math_score = "mean_scale_score", number_tested_math = "number_tested") %>% 
  select(dbn:mean_math_score)

#names(math_scores)
#str(math_scores)
#view(math_scores)

```

#adds building codes to math, ela and demo datasets
```{r, message=FALSE}
building_codes <- read_excel("./school_data/2017_water_lead.xlsx", sheet = "Building School Key", skip = 1) %>%
  janitor::clean_names() %>%
  mutate(school_name = toupper(school_name)) 


#view(building_codes)
#view(ela_scores)

ela_math <- full_join(math_scores, ela_scores) %>% 
  mutate(mean_ela_score = as.numeric(mean_ela_score),
         mean_math_score = as.numeric(mean_math_score)) %>%
  filter(grade == "All Grades") 

#view(ela_math)

codes_scores <- left_join(ela_math, building_codes, by = "school_name") %>% 
  rename(school_name_scores = school_name)
#view(codes_ela_scores)

codes_demo <- left_join(demo, building_codes, by = "school_name")
#view(codes_demo)
```

#merging files based on building code
```{r}
codes_demo_lead <- left_join(lead, codes_demo, by = "building_code") %>% 
  rename(school_name_lead = school_name)
#Adding lead data doubles our observations - something is wrong here. we have more schools tested for lead than tested for scores. E.g. buulding X has 2 schools in it, so 2 schools are assigned the lead data from testing that building. But only 1 school in the building was tested for math and ela. We needd to remove the school that was not tester by filtering so that school_name_scores == school_name_lead - the school name from the lead dataset equals the school name from the testing dataset.

#summary(codes_demo_lead)

#in some cases one school goes across multiple buildings. Summarized lead testing for each school.
codes_demo_lead_scores <- left_join(codes_demo_lead, codes_scores, by = "building_code")%>% filter(school_name_lead == school_name_scores) %>% group_by(year, school_name_scores) %>% 
  summarize(number_of_samples_tested = sum(number_of_samples_tested),
            number_of_elevated_samples = sum(number_of_elevated_samples),
            math_score = mean(mean_math_score),
            ela_score = mean(mean_ela_score)) %>% 
  mutate(percent_elevated = number_of_elevated_samples/number_of_samples_tested) %>% 
  ungroup() %>% rename(school_name=school_name_scores) 


final_data <- codes_demo_lead_scores %>% left_join(demo, by = "school_name") %>% drop_na()

#summary(final_data)

#numebrs make sense now - we have 2177 data points for test scores, and 2064 data points in our final data set. 
```

#hierarchical clustering 
https://uc-r.github.io/hc_clustering
```{r}
library(cluster)    # clustering algorithms
library(factoextra) # clustering visualization
#install.packages("factoextra")
library(dendextend) # for comparing two dendrograms
```

#HC analysis: 
#Outcome: lower test scores, math_score, ela_score
#Exposure: Elevated lead concentrations in school drinking water, percent_elevated

#creating the datasets for K means, heirarchical clustering analysis

```{r ses indicators}
dim(final_data)

# 1j Summary statistics on dataset

#summary(final_data)

# 1k Select the dataset

final_data_2017 =
  final_data %>%
  filter(year == "2017") %>%
  select(percent_white, percent_asian, percent_black, percent_hispanic, economic_need_index, percent_english_language_learners)

final_data_2016 =
  final_data %>%
  filter(year == "2016") %>%
  select(percent_white, percent_asian, percent_black, percent_hispanic, economic_need_index, percent_english_language_learners)

#summary(final_data_s)
#final_data_s <- as.data.frame(scale(final_data, center = TRUE))
```

#calculating Eucliean distance

```{r 2016}
#view(final_data)

# 6a.i Calculate euclidean distance between all combination of points

df.dist_2016 <- dist(final_data_2016, method = "euclidean")
df.dist_2017 <- dist(final_data_2017, method = "euclidean")

# 6a.ii Create hierarchial cluster solution
# hclust is an agglomerative clustering algorithm
# other functions are available in R 
# the method arguement determines the function used to compute distance between clusters
# We will just use the complete method for this lab. 

hc.complete_2016 <- hclust(df.dist_2016 , method = "complete")

hc.complete_2017 <- hclust(df.dist_2017, method = "complete")

# 6b Look at dendrogram

as.dendrogram(hc.complete_2016) %>% head()

as.dendrogram(hc.complete_2017) %>% head()
```

#plotting the cluster
```{r}
# hierarchial clusters are most clearly represented as visual trees

# 7a Plotting with Base R 

# 7a.i Extract the dendrogram from the HC solution 

dendro.complete_2016 <- as.dendrogram(hc.complete_2016)

dendro.complete_2017 <- as.dendrogram(hc.complete_2017)

# 7a.ii Plot 
# height indicates how dissimilar the two clusters are 
# i.e. the clusters fused at height = 22 
# are much more dissimilar 
# than the clusters fused at height = 4

dendro.complete_2016 %>% 
  plot(main = "Complete Linkage 2016 SES", ylab = "Height", leaflab = "none")

dendro.complete_2017 %>% 
  plot(main = "Complete Linkage 2017 SES", ylab = "Height", leaflab = "none")
```

#choosing the appropriate number of clusters 

```{r}
# 8a Plot branches of different cut heights
# we can use the color_branches() function to divide our data 
# into various numbers of clusters
# the dashed line (which we manually create with abline())
# indicates where the cut is taking place

#2016 
dendro.complete_2016  %>% 
  color_branches(k = 7) %>%
  plot(main = "Complete Linkage 2016 SES", k = 7, ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

dendro.complete_2016  %>% 
  color_branches(k = 20) %>%
  plot(main = "Complete Linkage 2016 SES", k = 20, ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

#2017
dendro.complete_2017  %>% #selected this one 
  color_branches(k = 7) %>%
  plot(main = "Complete Linkage 2017 SES", k = 7, ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

dendro.complete_2017 %>% #selected this one 
  color_branches(k = 20) %>%
  plot(main = "Complete Linkage 2017 SES", k = 20, ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

# 8c.i Cut the tree to have 5 clusters

hc.cluster.516 <- cutree(hc.complete_2016, 5)
# this shows us the number of members in each cluster
table(hc.cluster.516)

hc.cluster.517 <- cutree(hc.complete_2017, 5)
# this shows us the number of members in each cluster
table(hc.cluster.517)

# 8c.i Cut the tree to have 4 clusters

hc.cluster.416 <- cutree(hc.complete_2016, 4)
# this shows us the number of members in each cluster
table(hc.cluster.416)

hc.cluster.417 <- cutree(hc.complete_2017, 4)
# this shows us the number of members in each cluster
table(hc.cluster.417)

```


```{r}
# 8c.ii Include the cluster assignments in the original scaled data
#5 groups
df.hc.516 <- final_data_2016 %>% 
  mutate(hc.cluster.516 = hc.cluster.516) %>%
  rename("Asian" = "percent_asian",
         "White" = "percent_white",
         "Black" = "percent_black",
         "Hispanic" = "percent_hispanic",
         "Econ Need Index" = "economic_need_index",
         "ELL" = "percent_english_language_learners"
         )

df.hc.517 <- final_data_2017 %>%
  mutate(hc.cluster.517 = hc.cluster.517) %>%
  rename("Asian" = "percent_asian",
         "White" = "percent_white",
         "Black" = "percent_black",
         "Hispanic" = "percent_hispanic",
         "Econ Need Index" = "economic_need_index",
         "ELL" = "percent_english_language_learners"
         )

#4 groups
df.hc.416 <- final_data_2016 %>% 
  mutate(hc.cluster.416 = hc.cluster.416) %>%
  rename("Asian" = "percent_asian",
         "White" = "percent_white",
         "Black" = "percent_black",
         "Hispanic" = "percent_hispanic",
         "Econ Need Index" = "economic_need_index",
         "ELL" = "percent_english_language_learners"
         )

df.hc.417 <- final_data_2017 %>%
  mutate(hc.cluster.417 = hc.cluster.417) %>%
  rename("Asian" = "percent_asian",
         "White" = "percent_white",
         "Black" = "percent_black",
         "Hispanic" = "percent_hispanic",
         "Econ Need Index" = "economic_need_index",
         "ELL" = "percent_english_language_learners"
         )
# 8c.iv Compute mean concentration of each constituent within each cluster
#5 clusters
df.hc.mean.516 <- df.hc.516 %>% 
  group_by(hc.cluster.516) %>% 
  summarize_all(.funs = mean)

df.hc.mean.517 <- df.hc.517 %>% 
  group_by(hc.cluster.517) %>% 
  summarize_all(.funs = mean)

#4 clusters
df.hc.mean.416 <- df.hc.416 %>% 
  group_by(hc.cluster.416) %>% 
  summarize_all(.funs = mean)

df.hc.mean.417 <- df.hc.417 %>% 
  group_by(hc.cluster.417) %>% 
  summarize_all(.funs = mean)

# 8c.v Put data in long format
#5 groups
plot_means_h516 <- df.hc.mean.516 %>%
  gather(key = "group", value = "mean", -hc.cluster.516) 

plot_means_h517 <- df.hc.mean.517 %>%
  gather(key = "group", value = "mean", -hc.cluster.517)

#4 groups
plot_means_h416 <- df.hc.mean.416 %>%
  gather(key = "group", value = "mean", -hc.cluster.416) 

plot_means_h417 <- df.hc.mean.417 %>%
  gather(key = "group", value = "mean", -hc.cluster.417)

# 8c.vii.ii Plot
#5 groups
plot_5_16 <- ggplot(plot_means_h516, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, size = 0.2) +
  facet_wrap(~ hc.cluster.516, ncol = 5) +        # creates plots for each cluster
  theme_minimal() +  theme(axis.text = element_text(angle = 90), legend.position = "none") + labs(title = "SES Clusters 2016", x = "Group", y = "Mean %") +
  geom_bar(position = 'dodge', stat='identity') + geom_text(aes(label=round(mean, digits=1)), angle=90, label.size = 0.01, vjust=0.5) + scale_fill_manual(values = c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#984ea3"))

plot_5_17 <- ggplot(plot_means_h517, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, size = 0.2) +
  facet_wrap(~ hc.cluster.517, ncol = 5) +        # creates plots for each cluster
  theme_minimal() +  theme(axis.text = element_text(angle = 90), legend.position = "none") + labs(title = "SES Clusters 2017", x = "Group", y = "Mean %") +
  geom_bar(position = 'dodge', stat='identity') + geom_text(aes(label=round(mean, digits=1)), angle=90, label.size = 0.01, vjust=0.5) + scale_fill_manual(values = c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#984ea3"))

#combine both together
plot_5_16 + plot_5_17

```

#assign clusters based on math and ela scores!! 

```{r math and ela scores}
#create datasets
final_data_2017_score =
  final_data %>%
  filter(year == "2017") %>%
  select(ela_score, math_score)

final_data_2016_score =
  final_data %>%
  filter(year == "2016") %>%
  select(ela_score, math_score)


# 6a.i Calculate euclidean distance between all combination of points

df.dist_2016 <- dist(final_data_2016_score, method = "euclidean")

df.dist_2017 <- dist(final_data_2017_score, method = "euclidean")

# 6a.ii Create hierarchial cluster solution
# hclust is an agglomerative clustering algorithm
# other functions are available in R 
# the method arguement determines the function used to compute distance between clusters
# We will just use the complete method for this lab. 

hc.complete_2016 <- hclust(df.dist_2016 , method = "complete")
hc.complete_2017 <- hclust(df.dist_2017, method = "complete")


# 7a.i Extract the dendrogram from the HC solution 

dendro.complete_2016 <- as.dendrogram(hc.complete_2016)
dendro.complete_2017 <- as.dendrogram(hc.complete_2017)


# 7a.ii Plot 
# height indicates how dissimilar the two clusters are 
# i.e. the clusters fused at height = 22 
# are much more dissimilar 
# than the clusters fused at height = 4

dendro.complete_2016 %>% 
  plot(main = "Complete Linkage 2016 Scores", ylab = "Height", leaflab = "none")

dendro.complete_2017 %>% 
  plot(main = "Complete Linkage 2017 Scores", ylab = "Height", leaflab = "none")

# 8c.i Cut the tree to have 5 clusters

hc.cluster.216 <- cutree(hc.complete_2016, 3)
# this shows us the number of members in each cluster
table(hc.cluster.216)

hc.cluster.217 <- cutree(hc.complete_2017, 3)
# this shows us the number of members in each cluster
table(hc.cluster.217)

# 8c.ii Include the cluster assignments in the original scaled data

df.hc.216 <- final_data_2016_score %>% 
  mutate(hc.cluster.216 = hc.cluster.216) 

df.hc.217 <- final_data_2017_score %>% 
  mutate(hc.cluster.217 = hc.cluster.217) 

# 8c.iv Compute mean concentration of each constituent within each cluster

df.hc.mean.216 <- df.hc.216 %>% 
  group_by(hc.cluster.216) %>% 
  summarize_all(.funs = mean)

df.hc.mean.217 <- df.hc.217 %>% 
  group_by(hc.cluster.217) %>% 
  summarize_all(.funs = mean)

# 8c.v Put data in long format

plot_means_h216 <- df.hc.mean.216 %>%
  gather(key = "group", value = "mean", -hc.cluster.216) 
view(plot_means_h216)

plot_means_h217 <- df.hc.mean.217 %>%
  gather(key = "group", value = "mean", -hc.cluster.217) 
view(plot_means_h217)

# 8c.vii.ii Plot
ggplot(plot_means_h216, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, size = 0.2) +
  facet_wrap(~ hc.cluster.216) +        # creates plots for each cluster
  theme_bw() +  theme(axis.text = element_text(angle = 90)) + labs(title = "Scores 2016")

ggplot(plot_means_h217, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, size = 0.2) +
  facet_wrap(~ hc.cluster.217) +        # creates plots for each cluster
  theme_bw() +  theme(axis.text = element_text(angle = 90)) + labs(title = "Scores 2017")

#not much variation was observed in test scores but 3 clusters seems better than alternatives and is consistent across the two years
```


```{r percent elevated lead}
#create datasets
final_data_2017_lead =
  final_data %>%
  filter(year == "2017") %>%
  select(percent_elevated)

final_data_2016_lead =
  final_data %>%
  filter(year == "2016") %>%
  select(percent_elevated)


# 6a.i Calculate euclidean distance between all combination of points

df.dist_2016 <- dist(final_data_2016_lead, method = "euclidean")

df.dist_2017 <- dist(final_data_2017_lead, method = "euclidean")

# 6a.ii Create hierarchial cluster solution
# hclust is an agglomerative clustering algorithm
# other functions are available in R 
# the method arguement determines the function used to compute distance between clusters
# We will just use the complete method for this lab. 


hc.complete_2016 <- hclust(df.dist_2016, method = "complete")
hc.complete_2017 <- hclust(df.dist_2017, method = "complete")


# 7a.i Extract the dendrogram from the HC solution 

dendro.complete_2016 <- as.dendrogram(hc.complete_2016)
dendro.complete_2017 <- as.dendrogram(hc.complete_2017)

# 7a.ii Plot 
# height indicates how dissimilar the two clusters are 
# i.e. the clusters fused at height = 22 
# are much more dissimilar 
# than the clusters fused at height = 4

dendro.complete_2016 %>% 
  plot(main = "Complete Linkage 2016 Lead", ylab = "Height", leaflab = "none")

dendro.complete_2017 %>% 
  plot(main = "Complete Linkage 2017 Lead", ylab = "Height", leaflab = "none")

# 8a Plot branches of different cut heights

dendro.complete_2016  %>% 
  color_branches(k = 7) %>%
  plot(main = "Complete Linkage 2016 Lead, k = 7", ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

dendro.complete_2016 %>% #selected this one 
  color_branches(k = 20) %>%
  plot(main = "Complete Linkage 2016 Lead, k = 20", ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

dendro.complete_2017  %>% 
  color_branches(k = 7) %>%
  plot(main = "Complete Linkage 2017 Lead, k = 7", ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

dendro.complete_2017 %>% #selected this one 
  color_branches(k = 20) %>%
  plot(main = "Complete Linkage 2017 Lead, k = 20", ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)


# 8c.i Cut the tree to have 5 clusters

hc.cluster.216 <- cutree(hc.complete_2016, 3)
# this shows us the number of members in each cluster
table(hc.cluster.216)

hc.cluster.217 <- cutree(hc.complete_2017, 3)
# this shows us the number of members in each cluster
table(hc.cluster.217)

# 8c.ii Include the cluster assignments in the original scaled data

df.hc.216 <- final_data_2016_lead %>% 
  mutate(hc.cluster.216 = hc.cluster.216) 

df.hc.217 <- final_data_2017_lead %>% 
  mutate(hc.cluster.217 = hc.cluster.217) 

# 8c.iv Compute mean concentration of each constituent within each cluster

df.hc.mean.216 <- df.hc.216 %>% 
  group_by(hc.cluster.216) %>% 
  summarize_all(.funs = mean)

df.hc.mean.217 <- df.hc.217 %>% 
  group_by(hc.cluster.217) %>% 
  summarize_all(.funs = mean)

df.hc.mean.216
df.hc.mean.217
```


```{r percent elevated lead putting data in long format}
# 8c.v Put data in long format

plot_means_h216 <- df.hc.mean.216 %>%
  gather(key = "group", value = "mean", -hc.cluster.216) 
view(plot_means_h216)

plot_means_h217 <- df.hc.mean.217 %>%
  gather(key = "group", value = "mean", -hc.cluster.217) 
view(plot_means_h217)

# 8c.vii.ii Plot
ggplot(plot_means_h216, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, size = 0.2) +
  facet_wrap(~ hc.cluster.216) +        # creates plots for each cluster
  theme_bw() +  theme(axis.text = element_text(angle = 90)) + labs(title = "% Elevated Lead 2016")

ggplot(plot_means_h217, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, size = 0.2) +
  facet_wrap(~ hc.cluster.217) +        # creates plots for each cluster
  theme_bw() +  theme(axis.text = element_text(angle = 90)) + labs(title = "% Elevated Lead 2017")
```



```{r all exposures}
#create datasets
final_data_2017_all =
  final_data %>%
  filter(year == "2017") %>%
  select(percent_elevated, percent_white, percent_asian, percent_black, percent_hispanic, economic_need_index, percent_english_language_learners)

final_data_2016_all =
  final_data %>%
  filter(year == "2016") %>%
  select(percent_elevated, percent_white, percent_asian, percent_black, percent_hispanic, economic_need_index, percent_english_language_learners)


# 6a.i Calculate euclidean distance between all combination of points

df.dist_2016 <- dist(final_data_2016_all, method = "euclidean")

df.dist_2017 <- dist(final_data_2017_all, method = "euclidean")

# 6a.ii Create hierarchial cluster solution
# hclust is an agglomerative clustering algorithm
# other functions are available in R 
# the method arguement determines the function used to compute distance between clusters
# We will just use the complete method for this lab. 


hc.complete_2016 <- hclust(df.dist_2016, method = "complete")
hc.complete_2017 <- hclust(df.dist_2017, method = "complete")
```


```{r all exposures dendrograms}
# 7a.i Extract the dendrogram from the HC solution 

dendro.complete_2016 <- as.dendrogram(hc.complete_2016)
dendro.complete_2017 <- as.dendrogram(hc.complete_2017)

# 7a.ii Plot 
# height indicates how dissimilar the two clusters are 
# i.e. the clusters fused at height = 22 
# are much more dissimilar 
# than the clusters fused at height = 4

dendro.complete_2016 %>% 
  plot(main = "Complete Linkage 2016 All Exposures", ylab = "Height", leaflab = "none")

dendro.complete_2017 %>% 
  plot(main = "Complete Linkage 2017 All Exposures", ylab = "Height", leaflab = "none")

# 8a Plot branches of different cut heights

dendro.complete_2016  %>% 
  color_branches(k = 7) %>%
  plot(main = "Complete Linkage 2016 All Exposures, k = 7", ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

dendro.complete_2016 %>% #selected this one 
  color_branches(k = 20) %>%
  plot(main = "Complete Linkage 2016 All Exposures, k = 20", ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

dendro.complete_2017  %>% 
  color_branches(k = 7) %>%
  plot(main = "Complete Linkage 2017 All Exposures, k = 7", ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)

dendro.complete_2017 %>% #selected this one 
  color_branches(k = 20) %>%
  plot(main = "Complete Linkage 2017 All Exposures, k = 20", ylab = "Height", leaflab = "none") %>%
abline(h = 16, lty = 3)
```


```{r all exposures 5 clusters}
# 8c.i Cut the tree to have 5 clusters

hc.cluster.216 <- cutree(hc.complete_2016, 5)
# this shows us the number of members in each cluster
#table(hc.cluster.216)

hc.cluster.217 <- cutree(hc.complete_2017, 5)
# this shows us the number of members in each cluster
#table(hc.cluster.217)

# 8c.ii Include the cluster assignments in the original scaled data

df.hc.216 <- final_data_2016_all %>% 
  mutate(hc.cluster.216 = hc.cluster.216) %>%
  rename("Asian" = "percent_asian",
         "White" = "percent_white",
         "Black" = "percent_black",
         "Hispanic" = "percent_hispanic",
         "Econ Need Index" = "economic_need_index",
         "ELL" = "percent_english_language_learners",
         "Elevated Lead" = "percent_elevated"
         )

df.hc.217 <- final_data_2017_all %>% 
  mutate(hc.cluster.217 = hc.cluster.217) %>%
  rename("Asian" = "percent_asian",
         "White" = "percent_white",
         "Black" = "percent_black",
         "Hispanic" = "percent_hispanic",
         "Econ Need Index" = "economic_need_index",
         "ELL" = "percent_english_language_learners",
         "Elevated Lead" = "percent_elevated"
         )

# 8c.iv Compute mean concentration of each constituent within each cluster

df.hc.mean.216 <- df.hc.216 %>% 
  group_by(hc.cluster.216) %>% 
  summarize_all(.funs = mean)

df.hc.mean.217 <- df.hc.217 %>% 
  group_by(hc.cluster.217) %>% 
  summarize_all(.funs = mean)

# 8c.v Put data in long format

plot_means_h216 <- df.hc.mean.216 %>%
  gather(key = "group", value = "mean", -hc.cluster.216) 
#view(plot_means_h216)

plot_means_h217 <- df.hc.mean.217 %>%
  gather(key = "group", value = "mean", -hc.cluster.217) 
#view(plot_means_h217)

df.hc.mean.216
df.hc.mean.217
```


```{r all exposures 5 clusters plots}
# 8c.vii.ii Plot
a16 <- ggplot(plot_means_h216, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, size = 0.2) +
  facet_wrap(~ hc.cluster.216, ncol = 5) +        # creates plots for each cluster
  theme_bw() +  theme(axis.text = element_text(angle = 90), legend.position = "none") + labs(title = "All Exposures 2016", x = "Group", y = "Mean %") +
  geom_bar(position = 'dodge', stat='identity') + geom_text(aes(label=round(mean, digits=1)), angle=0, label.size = 0.10, vjust=0.5) + scale_fill_manual(values = c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fdbf6f", "#fb9a99", "#984ea3"))

a17 <- ggplot(plot_means_h217, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 0, size = 0.2) +
  facet_wrap(~ hc.cluster.217, ncol = 5) + # creates plots for each cluster
  theme_bw() +  theme(axis.text = element_text(angle = 90), legend.position = "none") + labs(title = "All Exposures 2017", x = "Group", y = "Mean %") +
  geom_bar(position = 'dodge', stat='identity') + geom_text(aes(label=round(mean, digits=1)), angle=0, label.size = 0.10, vjust=0.5) + scale_fill_manual(values = c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fdbf6f", "#fb9a99", "#984ea3"))

a16 + a17
```







